---
draft: true
title: Fraction of missing information
author: Edoardo Costantini
date: '2022-09-20'
slug: fmi
categories: ["missing values"]
bibliography: ../../resources/bibshelf.bib
toc-location: left
code-tools: true
toc-depth: 4
---

# Fraction of missing information

The **fraction of missing information** (FMI) is a measure used to quantify the impact of missing data on the estimation of a parameter $(Q)$.

> More specifically, it measures the inﬂation in the variance of the parameter estimate relative to what this variance would have been had all the data been observed.
>
> @savaleiRhemtulla:2012 [p478]

As a measure, it is the function of the proportion of missing data, the missing data mechanism, and the imputation procedure. When applying multiple imputation, the FMI is usually estimated as:

$$
\lambda = \frac{B + \frac{B}{m}}{T}
$$

where:

- $m$ is the number of imputations
- $B$, or between imputation variation, is the variance of the parameter estimate across the $m$ imputations
    $$
    B = \frac{1}{m - 1} \sum^{m}_{l=1} (\hat{Q}_{l} - \bar{Q}) (\hat{Q}_{l} - \bar{Q})'
    $$
    where:
    - $\hat{Q}_{l}$ is the estimate of the parameter of interest in the $l$-th imputed data
    - $\bar{Q} = \frac{1}{m} \sum^{m}_{l=1} \hat{Q}_{l}$
- $T$ is the total variation around the parameter estimate (as a result of the sum of the within and between imputation variation)
    $$
    T = \bar{U} + B + \frac{B}{m}
    $$
    where:
    - $\bar{U}$ is the average of the standard errors of $Q$ ($\bar{U}_{l}$) computed on the $m$ different datasets.
    $$
    \bar{U} = \frac{1}{m} \sum^{m}_{l = 1} \bar{U}_{l}
    $$

As the proportion of variation attributable to the missing data, the FMI can be interpreted as follow:

- $\lambda = 0$, the missing data do not add extra variation to parameter sampling variance;
- $\lambda = 1$, the missing data cause all of the variation in the parameter estimate;
- $\lambda > 0.5$, the influence of the imputation model on the final result is larger than that of the complete-data

### Relative increase in variance due to nonresponse

By replacing the denominator used in the estimation of $\lambda$ with $\bar{Q}$ we obtained a similar measure called **relative increase in variance due to nonresponse** ($riv$ or $r$):

$$
r = \frac{B + \frac{B}{m}}{\bar{Q}}
$$

This quantity is related to $\lambda$ by:

$$
r = \frac{\lambda}{1 - \lambda}
$$

### Fraction of information about Q missing due to nonresponse

Finally, a related measure is the **fraction of information about $Q$ missing due to nonresponse**:

$$
\gamma = \frac{r + \frac{2}{\nu + 3}}{1+r}
$$

where $\nu$ refers to the degrees of freedom, or the number of observations after accounting for the number of parameters in the model.
This quantity is related to $\lambda$ by:

$$
\gamma = \frac{\nu + 1}{\nu + 3} \lambda + \frac{2}{\nu + 3}
$$

The interpretation of $\gamma$ is the same as $\lambda$, but adjusted for the finite number of imputations.

## Degrees of freedom

The computation of the degrees of freedom is not the same as in a complete case analysis. There are actually a few different ways of computing $\nu$ for the multiple imputation of missing values.

- old formula for degrees of freedom [@rubin:1987, eq. 3.1.6]
    $$
    \nu_{old} = (m -1)(1+ \frac{1}{r^2}) = \frac{m-1}{\lambda^2}
    $$
    To think about these measure, consider:
    - if $\lambda \approx 1$, then $\nu_{old} \approx m - 1$, which occurs if close to all variation is due to nonresponse
    - if $\lambda \approx 0$, then $\nu_{old} \approx \infty$, which occurs if close to all variation is due to sampling variation
- observed data degrees of freedom
    $$
    \nu_{obs} = \frac{\nu_{com}+1}{\nu_{com}+3} \nu_{com}(1-\lambda)
    $$
    where $\nu_{com} = n - k$, the difference between the sample size and the number of parameters estimated.
- adjusted degrees of freedom
    $$
    \nu = \frac{\nu_{old}\nu_{obs}}{\nu_{old} + \nu_{obs}}
    $$
    This is the measure that should be used for statistical testing of estimates after applying multiple imputation. Note the following:
    - $\nu \le \nu_{com}$  always!
    - $\nu = \nu_{old}$ if $\nu_{com} = \infty$
    - $\nu = \nu_{com}$ if $\lambda = 0$
    - $\nu < 1$ you should refrain from testing due to lack of information in the data
    - $\nu = 0$ if $\lambda = 1$

# Learn by coding

Let's generate some fictitious data on which to try out the computations I described above.

## Data generation

First, let's load the `mice` package we will use to perform multiple imputations.

```{r}
# Environment set up -----------------------------------------------------------

# Load mice package
library(mice)

```

Then, let's sample three random variables from a multivariate normal distribution and use them to generate a dependent variable $y$.

```{r}
# Data generation --------------------------------------------------------------

# Set seed
set.seed(20230308)

# Define data generation model parameters
n <- 50     # sample size

# Sample X data from a multivariate normal distribution
X <- MASS::mvrnorm(
    n = n,
    mu = rep(0, 3),
    Sigma = matrix(c(
        1, .5, 0,
        .5, 1, 0,
        0, 0, 1
    ), nrow = 3)
)

# Data generation model parameter values
b0 <- 10    # intercept
b1 <- 5     # regression coefficients
b2 <- 5
b3 <- 0

# Sample Y from DGM
y <- b0 + b1 * X[, 1] + b2 * X[, 2] + b3 * X[, 3] + rnorm(n)

# Put together in a data frame
dt <- data.frame(y = y, X = X)

```

Finally, impose missing values by defining a vector of response probabilities based on the second column of $X$.

```{r}
# Compute the probabilities of nonresponse:
probs <- plogis(X[, 2])

# Return a logical nonresponse vector:
wy <- (as.logical(rbinom(n = n, size = 1, prob = probs)))

# Impose missingness
dt$y[wy] <- NA
```

We can now impute the data with the mice algorithm as implemented in the `mice()` R function. Let's obtain 20 imputations ($m = 20$). Every other argument is left to its default value as their specification is not particularly relevant for this demonstration.

```{r}
# Number of imputations (final data sets)
m <- 20

# Impute with Bayesian imputation (norm)
imp <- mice(dt, m = m, print = FALSE)

# Fit a model to the multiple imputations
fit.imp <- with(imp, lm(y ~ X.1 + X.2 + X.3))

# Pool statistics
pool.imp <- pool(fit.imp)

```

## FMI and variance ratios

To begin with, we need to compute the pooled estimates and measures of variability we need in the formulas of $\lambda$ and $r$.

### Estiamtes and variances

#### $\bar{Q}_l$

```{r}
# Compute manually ------------------------------------------------------------
# Notes: page and equation numbers refer to van Buuren 2018

# Extract the statistic of interest from every imputed data set
Q_bar_l <- sapply(
    fit.imp$analyses,
    function(x) {
        coef(x)["X.1"]
    }
)

```

#### $\bar{Q}$
```{r}
# > Compute Q_bar (p. 42, eq. 2.16) --------------------------------------------

# Obtain the pooled parameter
Q_bar <- mean(Q_bar_l)

```

#### $\bar{U}_{l}$ and $\bar{U}$
```{r}
# > Compute U_bar (p. 43, eq. 2.18) -------------------------------------------

# Extract vector of standard errors
U_bar_l <- sapply(
    fit.imp$analyses,
    function(x) {
        (coef(summary(x))["X.1", "Std. Error"])^2
    }
)

# Obtain the pooled parameter
U_bar <- mean(U_bar_l)

```

#### $B$
```{r}
# > Compute B (p. 43, eq. 2.19) -----------------------------------------------

B <- sum((Q_bar_l - Q_bar)^2) / (m - 1)

```

#### $T$
```{r}
# > Compute T (p. 43, eq. 2.20) -----------------------------------------------

T <- U_bar + B + B/m

```

### Variance ratios

We can now compute all of the variance ratios discussed.
The proportion of variation attributable to the missing data ($\lambda$) can be computed with the following code:

```{r}
# > Lambda (p. 46, eq. 2.24) --------------------------------------------------

lambda <- (B + B/m) / T

```

Note that, for infinite imputations $m \rightarrow \infty$, $\lambda$ is equal to the ratio $B / T$, which is the proportion of the total parameter estimate variance that is attributable to between imputation variance.

```{r, eval = FALSE}
# Conceptual lambda
B / T

```

The computation of $r$ is:

```{r}
# > Relative increase in variance due to nonresponse (p. 47, eq. 2.25) --------

r <- (B + B / m) / U_bar

```

and note again the conceptual meaning:

```{r, eval = FALSE}
# Conceptual meaning
B / U_bar

```

and the relationship between $r$ and $\lambda$ is as described

```{r}
# r relation to lambda
r == lambda / (1-lambda)

```

Let's now focus on the degrees of freedom.

```{r}
# > Degrees of freedom (old) --------------------------------------------------
#   Interpretation: number of observations after accounting for the number of 
#                   parameters in the model.

    nu_old <- (m - 1) * (1 + 1 / r^2)
    nu_old <- (m - 1) / lambda^2

# > Degrees of freedom (com) --------------------------------------------------
#   Interpretation: degrees of freedom of Q_bar in the hypothetically complete
#                   data

    # number of parameters
    k <- 1 + 3

    # Compute degrees of freedom
    nu_com <- n - k

# > Degrees of freedom (obs) --------------------------------------------------
#   Interpretation: estimated observed data degrees of freedom accounting for 
#                   missing information

    nu_obs <- (nu_com + 1) / (nu_com + 3) * nu_com * (1 - lambda)

# > Degrees of freedom (adjusted) ---------------------------------------------
#   Interpretation: adjusted degrees of freedom to be used for testing in
#                   multiple imputation

    nu <- nu_old * nu_obs / (nu_old + nu_obs)
```

With all of the degrees of freedom computed, we can now calculate the Fraction of information missing due to nonresponse

```{r}
# > Fraction of information missing due to nonresponse (p. 47, eq. 2.26) ------
#   Interpretation: proportion of variation attributable to the missing data
#                   -> ADJUSTED for the finite number of imputations

    gamma <- (r + 2/(nu + 3)) / (1 + r)

    # Relation to lambda
    # 1. Interpretation: same + ADJUSTED for a finite number of imputations
    # 2. Computation:
    c(
        gamma,
        (nu + 1) / (nu + 3) * lambda + 2 / (nu + 3)
    )
    # 3. Use in the literature: used interchangeably, but only for large nu 
    #                           they are the same.

# > Compare with mice ---------------------------------------------------------

    # Results for B1
    manual <- c(
        Q_bar = Q_bar, U_bar = U_bar, B = B, T = T,
        nu_com = nu_com, nu = nu,
        r = r, lambda = lambda, gamma = gamma
    )

    # Mice results
    mice_comp <- pool.imp$pooled[2, -(1:2)]

    # Put together
    rbind(manual = manual, mice_comp = mice_comp)
```


# TL;DR, just give me the code!
```{r TLDR, ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}
```