---
title: Understanding quantiles
author: Edoardo Costantini
date: '2022-09-06'
slug: quantiles
categories: ["Statistics", "Distributions"]
---

# Quantiles, percentiles, and quartiles

Descriptive statistics are a fundamental tools for a data analysis.
Their purpose is to describe and summarize data.
Quantitative variables such as weight, age, and income can be described by distributions (e.g., normal distribution) with measures of **center** (the typical value of a variable) and **variability** (spread around the center).
The mean is a measure of center.
The standard deviation is a measure of variability.

There are special descriptive statistics that help us describe simultaneously center and spread of a variable's distribution.
These measures are often known as measures of **positions**.
In general, they tell us the point of the distribution of a variable at which a given percentage of the data falls below or above that point.
The minimum and maximum values of a variable are measures of positions defining the point at which no data, or all data, fall below it, respectively.
The median is a measure of position that defines the point at which half of the data falls below it (and above it).
The median is a special case of the measure of position called **percentile**.
The p-th percentile is the point such that p\% of the observations fall below or at that point and (100 - p)% fall above it.
The 50-th percentile is the median.

**Quartiles** are commonly used percentiles.
The first quartile is the 25-th percentile, the value of the variable leaving to its left 25% of the distribution.
The third quartile is the 75-th percentile, the value of the variable leaving to its left 75% of the distribution.
You can think of quartiles as diving the probability distribution in 4 intervals with equal probabilities (e.g., area under the probability density function).
Similarly, you can think of percentiles as diving the probability distribution in 100 intervals with equal probabilities.

```{r quartile-pic, warning = FALSE, message = FALSE, echo=FALSE, fig.width=10, fig.height = 8}

# Set a seed

set.seed(20220906)

# Sample from a normal distribution

age <- rnorm(1e5, mean = 27, sd = 2)

# Define the 1st and 2nd quartile

quartiles <- quantile(age, probs = c(.25, .75))

# Plot density distribution

plot(density(age),
     main = "Quartiles for the probability distribution of age",
     xlab = NA)

# Costum x ticks
axis(side = 1, at = c(27, round(quartiles, 1)), labels = TRUE)

# Add points for quantiles
points(c(quartiles, median(age)),
        y = rep(0, length(quartiles)+1))
points(c(quartiles, median(age)),
       y = dnorm(c(quartiles, median(age)), mean = mean(age), sd = sd(age)))

# Add segments to devide plot
segments(x0 = quartiles[1], y0 = 0,
         x1 = quartiles[1], y1 = dnorm(quartiles[1],
                                       mean = mean(age), sd = sd(age)))
segments(x0 = median(age), y0 = 0,
         x1 = median(age), y1 = max(dnorm(age,
                                       mean = mean(age), sd = sd(age))))
segments(x0 = quartiles[2], y0 = 0,
         x1 = quartiles[2], y1 = dnorm(quartiles[2],
                                       mean = mean(age), sd = sd(age)))

# Add quartile labels
text(x = quartiles[1],
     y = -.005,
     "1st quartile")
text(x = quartiles[2],
     y = -.005,
     "3rd quartile")

# Add percentage under the curve labels
text(x = c(24, 30, 26.3, 27.7),
     y = c(.03, .03, .06, .06),
     "25 %")

```


We use the word **quantile** to describe the general measure of position that divides the probability distribution in intervals with equal probabilities.
Percentiles divide the probability distribution of a variable in 100 intervals.
Deciles divide it in 10 intervals, and quartiles in four.
As such, percentiles, deciles, quartiles, and the median are all special cases of quantiles.

# TL;DR, just give me the code!
```{r TLDR, ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}
```