---
draft: true # true
title: Understanding t-tests
author: Edoardo Costantini
date: '2022-09-06'
slug: t-test
categories: ["statistics", "inference", "distributions"]
---

# Introduction

A t-test is an inferential statistic used to draw conclusions based on comparisons between means.
Like any statistical test, it weights the estimated effect by the precision of the estimate.

$$
\text{test statistic} = \frac{\text{estimate} − \text{Null Hypothesized Value}}{\text{Variability}}
$$

The shape this formula takes depends on the type of test we want to perform.
T-tests are used to compare means, so the value of `estimate` and the `Null-Hypothesized Value` will be means and the variability will be the standard error of the mean.

## 1-sample t-test

The simplest form of t-test is what is known as a 1-

## 2-sample t-test

### Unequal group sizes

### Unequal variances between groups

```{r quartile-pic, warning = FALSE, message = FALSE, echo=FALSE, fig.width=10, fig.height = 8}

# shinyApp(
#     ui = fluidPage(
#       numericInput("n", "n", 1),
#       plotOutput("plot")
#     ),
#     server = function(input, output) {
#       output$plot <- renderPlot( plot(head(cars, input$n)) )
#     }
#   )


```

## TL;DR, just give me the code!
```{r TLDR, ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}
```